# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
First, unzip the .csv file and read it into R.  
```{r}
library("reshape2")
unzip("activity.zip", files = "activity.csv")
data <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
First sum the number of steps per day.
```{r}
datamelt <- melt(data, id = "date", measure.vars = "steps")
sumSteps <- dcast(datamelt, date ~ variable, sum)
```
Plot
```{r}
hist(sumSteps$steps, xlab = "Steps", main = "Total Steps Taken")
```
  
  
Find the mean and median
```{r}
mean(sumSteps$steps, na.rm = TRUE)
median(sumSteps$steps, na.rm = TRUE)
```


## What is the average daily activity pattern?

```{r}
datamelt2 <- melt(data, id = "interval", measure.vars = "steps", na.rm = TRUE)
avgSteps <- dcast(datamelt2, interval ~ variable, mean)
```

Make the plot
  
```{r}
plot(avgSteps, type = "l")
```
  
Find the interval on average across all days contains the maximum number of steps
```{r}
avgSteps[avgSteps$steps==max(avgSteps$steps),]
```


## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
